\Large \bf Uncovering Distributed System Bugs during Testing (not Production!)

### 2015-09-18 overview.tex
many updates in section 3 to address Ally's comments

\subsection{Specifications}
\label{sec:bg:bugs}
In  this section, we describe how safety and liveness specifications are expressed in \psharp.

{\bf Safety}.
In addition to the usual assertions that can be used to write safety properties that are local to a machine, \psharp also provides a way to specify global assertions by using a \emph{safety monitor}~\cite{desai2015building}, which is a special \psharp machine that can receive, but not send, events. A safety monitor maintain local state that is modified in response to events received from regular (i.e. non-monitor) machines. This local state is used to maintain a history of the computation that is relevant to the property being specified. An erroneous global behavior is flagged via an assertion on the private state of the safety monitor.

To declare a monitor, the programmer inherits from the \psharp \texttt{Monitor} class. Monitors in \psharp are singleton instances and, thus, a regular machine does not need a reference to a monitor to send an event to it.
A regular machine can send an event to a monitor by calling \texttt{Monitor<M>(e, p)}, where the parameter \texttt{e} is the event being send together with an optional payload \texttt{p}.

{\bf Liveness}.
Liveness property specifications are more difficult to encode since they are intended to capture program progress
over infinite executions.
Violation of a liveness property can only be demonstrated by an infinite execution.
Usually, a liveness property is specified via a temporal logic formula~\cite{Pnueli1977,lamport1994temporal}.
We take a different approach and allow the programmer to write a \emph{liveness monitor}.
Similar to a safety monitor, a liveness monitor receives events but does not send them.
Unlike a safety monitor, a liveness monitor~\cite{desai2015building}
contains three types of states---\emph{ordinary}, \emph{hot}, and \emph{cold}.
A hot state denotes a point in the execution where progress is required but has not happened yet,
e.g., a node has failed but a new one has not come up yet.
A liveness monitor enters a hot state upon receiving notification of an event that requires the system
to make progress and leaves the hot state upon receiving another event notifying it of progress.
An infinite execution is erroneous if the liveness monitor is in a hot state infinitely often but in a cold state only finitely often.
Our liveness monitors can encode arbitrary temporal logic properties.

@@

The \psharp runtime is a lightweight layer build on top of the Task Parallel Library (TPL) of .NET that implements the semantics of \psharp: creating state machines, executing them concurrently using the default task scheduler of TPL, and sending events and enqueueing them in the appropriate machines. A key capability of the \psharp runtime is that it can execute in bug-finding mode for systematically testing a \psharp program to find bugs, such as assertion violations and uncaught exceptions. We now give an overview of how this works; more details can be found in a previous paper~\cite{deligiannis2015psharp}.

When the \psharp runtime runs in bug-finding mode, an embedded \emph{systematic concurrency testing engine}
~\cite{godefroid1997verisoft, musuvathi2008finding, emmi2011delay} captures and takes control of all sources
of non-determinism that are \emph{known} to the \psharp runtime.
In particular, the runtime is aware of nondeterminism due to interleaving of event handlers in different machines.
Each send operation to an ordinary (non-monitor) machine and each create operation of an ordinary (non-monitor) machine creates an
interleaving point where the runtime makes a scheduling decision.
The engine will repeatedly execute a program from start to completion, each time exploring a potentially different set of interleavings,
until it either reaches a bound (in number of iterations or time), or it hits a property violation.
The testing is fully automatic, has no false-positives (assuming an accurate environmental model),
and can reproduce found bugs by replaying buggy schedules.
After a bug is found, the engine generates a trace that represents the buggy schedule.
In contrast to typical logs generating during production, this log provides a global order of all communication events
and is easy to debug.

We have implemented two schedulers inside the \psharp systematic testing engine:
a \emph{random} scheduler that makes each nondeterministic choice randomly
and \emph{PCT} scheduler~\cite{burckhardt2010pct}.
It is straightforward to create a new scheduler by implementing the \texttt{ISchedulingStrategy} interface~\cite{desai2015systematic}
exposed by the \psharp libraries.
The interface exposes callbacks that are invoked by the \psharp runtime for taking decisions regarding which machine to schedule next,
and can be used for developing both generic and application-specific schedulers.

@@

Vanilla \psharp is able to capture and take control of the \emph{inter-machine concurrency} due to message passing, but is unable to systematically explore any interleavings due to \emph{intra-machine concurrency} (e.g. async/await or TPL). This is problematic as nowadays, with multicore machines being a commodity, programmers tend to write multithreaded code to exploit shared memory architectures and increase the performance of individual components of a distributed system.

As an example, the Live Azure Migration system uses the \texttt{async} and \texttt{await} \csharp 5.0 language primitives. Asynchronous code using async/await is more readable because it looks like traditional procedural code, but it is translated by the compiler to an event-driven state machine that is built on top of TPL to achieve performance. The key idea behind async/await is that a method declared as \texttt{async} can use internally the \texttt{await} keyword which allows the thread executing the method to wait on a TPL task, or any other \emph{awaitable} object, \emph{without} blocking. This is achieved as follows. When an \texttt{await} statement is executed, the code following the await is wrapped as a TPL task \emph{continuation} and the method returns to the caller. This continuation executes when the awaitable object has completed.

Developing a fully automatic and universal approach to handle intra-machine concurrency in .NET is very challenging, because there are many APIs that can be used for concurrency and synchronization (e.g. \texttt{System.Threading}, TPL, async/await, locks, semaphores) and each one has its own complexities. Although developers are willing to model the top-level message passing communication using \psharp, they are resistant in modeling the low-level threading and synchronization methods as this would be a very invasive procedure and such refactoring is unlikely to scale for legacy code.

To test our case studies, we decided to focus our efforts in handling async/await as providing a robust solution for a subset of the .NET threading APIs is more feasible than trying to handle arbitrary threading and synchronization. Our solution involves using a custom task scheduler whenever the bug-finding mode of the \psharp runtime is enabled, instead of the default TPL task scheduler. The \psharp task scheduler inherits from the \texttt{TaskScheduler} class, a low-level API that is responsible for enqueueing TPL tasks into threads.

Our approach works as follows. We start the task of the \emph{root} \psharp machine in our custom task scheduler. As long as any child tasks spawned by the root machine task do not explicitly start in another scheduler (including the default TPL scheduler), then they will be scheduled for execution in our custom task scheduler. Our custom scheduler intercepts the call to enqueue a task, creates a special machine called \texttt{TaskMachine} and \emph{wraps} the enqueued task inside this machine. The constructor of a \texttt{TaskMachine} takes as an argument a TPL task and stores it in an field. The \texttt{TaskMachine} has only a single state and when it is scheduled for execution by the \psharp systematic testing scheduler it will start executing the wrapped task and then immediately wait on its completion. This forces the lifetime of the task to become the lifetime of the machine and vice versa. Thus, when the \psharp systematic testing scheduler schedules the \texttt{TaskMachine}, it will also schedule the corresponding task. This means that when a \psharp program makes a call to an \texttt{async} method, then the task created in the backend will be automatically wrapped and scheduled.

@@

Regarding support for handling non-async/await concurrency primitives, we are currently investigate feasible approaches.

>>>>

\subsection{Specifications}
\label{sec:bg:bugs}

In  this section, we describe how safety and liveness specifications are expressed in \psharp.

\textbf{Safety}.
In addition to the usual assertions that can be used to write safety properties that are local to a machine, \psharp also provides a way to specify global assertions by using a \emph{safety monitor}~\cite{desai2015building}, which is a special machine that can receive, but not send, events. A safety monitor maintain local state that is modified in response to events received from ordinary (i.e. non-monitor) machines. This local state is used to maintain a history of the computation that is relevant to the property being specified. An erroneous global behavior is flagged via an assertion on the private state of the safety monitor.

To declare a monitor, the programmer inherits from the \psharp \texttt{Monitor} class. Monitors in \psharp are singleton instances, which means that an ordinary machine does not need a reference to a monitor to send it an event. An ordinary machine can send an event to a monitor by calling \texttt{Monitor<M>(e, p)}, where the parameter \texttt{e} is the event being send together with an optional payload \texttt{p}.

\textbf{Liveness}.
Liveness property specifications are more difficult to encode since they are intended to capture program progress
over infinite executions.
Violation of a liveness property can only be demonstrated by an infinite execution.
Usually, a liveness property is specified via a temporal logic formula~\cite{Pnueli1977,lamport1994temporal}.
We take a different approach and allow the programmer to write a \emph{liveness monitor}.
Similar to a safety monitor, a liveness monitor receives events but does not send them.
Unlike a safety monitor, a liveness monitor~\cite{desai2015building}
contains three types of states---\emph{ordinary}, \emph{hot}, and \emph{cold}.
A hot state denotes a point in the execution where progress is required but has not happened yet,
e.g., a node has failed but a new one has not come up yet.
A liveness monitor enters a hot state upon receiving notification of an event that requires the system
to make progress and leaves the hot state upon receiving another event notifying it of progress.
An infinite execution is erroneous if the liveness monitor is in a hot state infinitely often but in a cold state only finitely often.
Our liveness monitors can encode arbitrary temporal logic properties.

@@

The \psharp runtime is a lightweight layer build on top of the Task Parallel Library (TPL) of .NET that implements the semantics of \psharp: creating state machines, executing them concurrently using the default task scheduler of TPL, and sending events and enqueueing them in the appropriate machines. A key capability of the \psharp runtime is that it can execute in bug-finding mode, which systematically tests a \psharp program to find bugs using the safety and liveness monitors that were discussed in Section~\ref{sec:bg:bugs}.

When the \psharp runtime executes in bug-finding mode, an embedded \emph{systematic concurrency testing engine}~\cite{godefroid1997verisoft, musuvathi2008finding, emmi2011delay} captures and takes control of all sources of non-determinism that are \emph{known} to the \psharp runtime. In particular, the runtime is aware of nondeterminism due to the interleaving of event handlers in different machines. Each send operation to an ordinary (non-monitor) machine, and each create operation of an ordinary machine, creates an interleaving point where the runtime makes a scheduling decision. The runtime is also aware of all nondeterministic events (e.g. failures) that were captured during the modeling process.

The \psharp systematic testing engine will repeatedly execute a program from start to completion, each time exploring a different set of scheduling decisions, until it either reaches a bound (in number of iterations or time), or it hits a property violation. This testing process is fully automatic, has no false-positives (assuming an accurate environmental model), and can reproduce found bugs by replaying buggy schedules. After a bug is found, the engine generates a trace that represents the buggy schedule. In contrast to logs typically generating during production, this log provides a global order of all communication events and, thus, is easier to debug.

We have implemented two different schedulers inside the \psharp systematic testing engine: a \emph{random} scheduler that makes each nondeterministic choice randomly and a \emph{probabilistic concurrency testing}~\cite{burckhardt2010pct} (PCT) scheduler. It is straightforward to create a new scheduler by implementing the \texttt{ISchedulingStrategy} interface~\cite{desai2015systematic} exposed by the \psharp libraries. The interface exposes callbacks that are invoked by the \psharp runtime for taking decisions regarding which machine to schedule next, and can be used for developing both generic and application-specific schedulers.

@@

The original implementation of \psharp is able to capture and take control of the \emph{inter-machine concurrency} due to message passing, but is unable to systematically explore any interleavings due to \emph{intra-machine concurrency} (e.g. async/await, which we discuss below). This is problematic since programmers often write multithreaded code that exploits shared memory multicore architectures, to increase the performance of individual components of a distributed system.

As an example, the Live Azure Migration system uses the \texttt{async} and \texttt{await} \csharp 5.0 language primitives. Asynchronous code using async/await is more readable than older asynchronous programming libraries in .NET, because it looks like traditional procedural code, but it is translated by the compiler to an event-driven state machine that is built on top of TPL to achieve performance. The key idea behind async/await is that a method declared as \texttt{async} can use internally the \texttt{await} keyword which allows the thread executing the method to wait on a TPL task, or any other \emph{awaitable} object, \emph{without} blocking. This is achieved as follows. When an \texttt{await} statement is executed, the code following the await is wrapped as a TPL task \emph{continuation} and the method returns to the caller. This continuation executes when the awaitable object has completed.

Developing a fully automatic and universal approach to handle intra-machine concurrency in .NET is very challenging, because there are many APIs that can be used for concurrency and synchronization (e.g. \texttt{System.Threading}, TPL, async/await, locks, semaphores) and each one has its own complexities. Although developers are willing to model the top-level message passing communication using \psharp, in our experience they are resistant to modeling the low-level threading and synchronization methods as this would be a very invasive procedure and such refactoring is unlikely to scale for legacy code.

To test our case studies, we decided to focus our efforts in handling async/await as providing a robust solution for a subset of the .NET threading APIs is more feasible than trying to handle arbitrary threading and synchronization. By choosing to focus on async/await, we cannot currently systematically test programs that use other mechanisms for intra-machine concurrency.  We can still apply \psharp on them, but we lose the benefit of systematic testing. In future work, we plan to investigate support for non-async/await concurrency primitives.

Our solution involves using a custom task scheduler whenever the bug-finding mode of the \psharp runtime is enabled, instead of the default TPL task scheduler. The \psharp task scheduler inherits from the \texttt{TaskScheduler} class, a low-level API that is responsible for enqueueing TPL tasks into threads. Our approach works as follows. We start the task of the \emph{root} \psharp machine in our custom task scheduler. As long as any child tasks spawned by the root machine task do not explicitly start in another scheduler (including the default TPL scheduler), then they will be scheduled for execution in our custom task scheduler. Our custom scheduler intercepts the call to enqueue a task, creates a special machine (only internally available to \psharp) called \texttt{TaskMachine} and \emph{wraps} the enqueued task inside this machine. The constructor of a \texttt{TaskMachine} takes as an argument a TPL task and stores it in an field. The \texttt{TaskMachine} has only a single state and when it is scheduled for execution by the \psharp systematic testing scheduler it starts executing the wrapped task and then immediately waits on its completion. This forces the lifetime of the task to become the lifetime of the machine and vice versa. Thus, when the \psharp systematic testing scheduler schedules the \texttt{TaskMachine}, it will also schedule the corresponding task. This means that when a \psharp program makes a call to an \texttt{async} method, then the task created in the backend will be automatically wrapped and scheduled.

@@
